Hydra Battles and Cie

### 2021-07-26 doc/chapter-primrec.tex
chapter 9 with alectryon: first draft

\begin{Coqsrc}
Lemma majorSucc : majorizedPR  succFunc Ack.

Lemma majorZero : majorizedPR  zeroFunc Ack.

Lemma majorProjection (n m:nat)(H: m < n):
  majorizedPR (projFunc n m H) Ack.
\end{Coqsrc}

The rest of the cases are proved within a mutual  induction.

@@

\begin{Coqsrc}
Lemma majorAnyPR:  forall n (x: PrimRec n),  majorizedPR  x Ack.
Proof.
  intros n x; induction x using PrimRec_PrimRecs_ind with
                  (P0 := fun n m y => majorizedSPR  y Ack).
  - apply majorSucc.
  - apply majorZero.
  - apply majorProjection. 
  \end{Coqsrc}

  \begin{Coqsrc}
  - (** function composition *)
\end{Coqsrc}
\begin{Coqanswer}
1 subgoal (ID 265)

  n, m : nat
  g : PrimRecs n m
  x : PrimRec m
  IHx : majorizedSPR g Ack
  IHx0 : majorizedPR x Ack
  ============================
  majorizedPR (composeFunc n m g x) Ack
\end{Coqanswer}

\begin{Coqsrc}
  ...
 - (** primitive recursion *)
\end{Coqsrc}

\begin{Coqanswer}
1 subgoal (ID 265)

  n : nat
  x1 : PrimRec n
  x2 : PrimRec (S (S n))
  IHx1 : majorizedPR x1 Ack
  IHx2 : majorizedPR x2 Ack
  ============================
  majorizedPR (primRecFunc n x1 x2) Ack
\end{Coqanswer}

\begin{Coqsrc}
 assert (L1 : forall i (v: t nat n) ,
               v_apply f (i::v)  <= Ack q (i + max_v v)).
    { induction i.
      ...
    }
    ...
-  (** empty list of functions *)
\end{Coqsrc}

\begin{Coqanswer}
1 subgoal (ID 266)

  n : nat
  ============================
  majorizedSPR (PRnil n) Ack
\end{Coqanswer}

\begin{Coqsrc}
  ...
- (** non-empty list of functions *)
\end{Coqsrc}

\begin{Coqanswer}
1 subgoal (ID 273)

  n, m : nat
  x : PrimRec n
  p : PrimRecs n m
  IHx : majorizedPR x Ack
  IHx0 : majorizedSPR p Ack
  ============================
  majorizedSPR (PRcons n m x p) Ack
\end{Coqanswer}

\begin{Coqsrc}
 ...
Qed.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma majorPR2 (f: naryFunc 2)(Hf : isPR 2 f)
  : exists (n:nat), forall x y,  f x y <= Ack n (max x  y).
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma Ack_strict_mono_l : forall n m p, n < m ->
                                        Ack n (S p) < Ack m (S p).
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma majorPR2_strict (f: naryFunc 2)(Hf : isPR 2 f):
    exists (n:nat),
    forall x y, 2 <= x -> 2 <= y -> f x y < Ack n (max x  y).
\end{Coqsrc}

@@

\begin{Coqsrc}
Section Impossibility_Proof.

  Hypothesis HAck : isPR 2 Ack.

  Lemma Ack_not_PR : False.
  Proof.
    destruct (majorPR2_strict Ack HAck) as [m Hm];
    pose (X := max 2 m); specialize (Hm X X).
    rewrite max_idempotent in Hm; 
    assert (Ack m X <= Ack X X) by (apply Ack_mono_l; lia).
    lia.
  Qed.

End Impossibility_Proof.
\end{Coqsrc}

\begin{remark}
It is easy to prove that any unary function which dominatates \texttt{fun n => Ack n n} fails to be primitive recursive. We use an instance of \texttt{majorAnyPR} for unary functions.

@@

\emph{From \href{../theories/html/hydras.MoreAck.AckNotPR.html}{MoreAck.AckNotPR}}.
\begin{Coqsrc}
Lemma majorPR1  (f: naryFunc 1)(Hf : isPR 1 f)
  : exists (n:nat), forall x, f x <= Ack n x.
(* ... *)
\end{Coqsrc}

@@

\begin{Coqsrc}
Section dom_AckNotPR.

  Variable f : nat -> nat.
  Hypothesis Hf : dominates f (fun n => Ack n n).

 Lemma dom_AckNotPR: isPR 1 f -> False.
  Proof.
    intros H;  destruct Hf as [N HN].
    destruct  (majorPR1 _ H) as [M HM].
    pose (X := Max.max N M).
    specialize (HN X  (Max.le_max_l N M)).
    specialize (HM X);
      assert (Ack M X <= Ack X X) by (apply Ack_mono_l; subst; lia).
    lia.
  Qed.

End dom_AckNotPR.
\end{Coqsrc}
\end{remark}

@@

\begin{Coqsrc}
Definition l_std alpha k := (L_ alpha (S k) - k)%nat.

Lemma l_std_ok : forall alpha : E0,
    alpha <> Zero ->
    forall k : nat,
      1 <= k -> battle_length standard k (iota (cnf alpha))
                              (l_std alpha k).
(* ... *)
\end{Coqsrc}

@@

\begin{Coqsrc}
Section battle_lenght_notPR.

  (** We assume that the function with computes the length 
      of standard battles is primitive recursive *)

  Hypothesis H: forall alpha, isPR 1 (l_std alpha). 
\end{Coqsrc}

@@

\begin{Coqsrc}
Let alpha := Phi0 omega%e0.
Let h := iota (cnf alpha).
\end{Coqsrc}

@@

\begin{Coqsrc}
Let m k := L_ alpha (S k).

Remark m_eqn : forall k, m k = (l_std alpha k + k)%nat.
(* ... *)
\end{Coqsrc}

@@

\begin{Coqsrc}
Remark mIsPR : isPR 1 m.
(* ... *)
\end{Coqsrc}

@@

In \href{../theories/html/hydras.Epsilon0.F_alpha.html}{Epsilon0.F\_alpha}, we prove a relation between the $F$ and $H'$ functional. For any $\alpha$ and $k>0$,
$H'_{\omega^\alpha}(k)\geq F_\alpha(k)$.

\begin{Coqsrc}
H'_F : forall (alpha : E0) (n : nat), 
     F_ alpha (S n) <= H'_ (Phi0 alpha) (S n) 
\end{Coqsrc}

Our proof of this lemma is not trivial at all. One of it sub-goals is the following one:

\begin{Coqsrc}
  alpha : E0
  IHalpha : forall beta : E0, beta o< alpha -> P beta
  Halpha : Limitb alpha
  n : nat
  ============================
  H'_ (Phi0 (CanonS alpha n)) (S n) <= 
  H'_ (Phi0 (CanonS alpha (S n))) (S n)
\end{Coqsrc}

In mathematical notation: $H'_{\omega^{\canonseq{\alpha}{n}}}(n+1) \leq
H'_{\omega^{\canonseq{\alpha}{n+1}}}(n+1)$.

But there exists no lemma saying that, if 
$\beta\leq \alpha$, then $H'_\beta(k)\leq H'_\alpha(k)$, for any $\alpha$ and $\beta$. For instance, 
$H'_{42}(3)=45> H'_\omega(3)=7$.

Looking for lemmas of the form $H'_\beta(k)\leq H'_\alpha(k)$, we find this one (from our library
\href{../theories/html/hydras.Epsilon0.Hprime.html}{Epsilon0.Hprime}):

\begin{Coqanswer}
H'_restricted_mono_l : 
    forall (alpha beta : E0) (n : nat), 
      Canon_plus n alpha beta -> 
      H'_ beta n <= H'_ alpha n.
\end{Coqanswer}

Thus, it remains to prove that 
there exists a path from ${\omega^{\canonseq{\alpha}{n+1}}}$
to ${\omega^{\canonseq{\alpha}{n}}}$ composed of 
$n+1$-steps.

Fortunately, the Ketonen-Solovay machinery contains three lemmas which help us to build such a path.

\begin{Coqanswer}
KS_thm_2_4_lemma5 :
  forall [i : nat] [alpha beta : T1],
  const_pathS i alpha beta ->
  nf alpha -> alpha <> zero -> 
  const_pathS i (phi0 alpha) (phi0 beta)

KS_thm_2_4 :
  forall [lambda : T1], nf lambda ->limitb lambda ->
  forall i j : nat, i < j -> 
   const_pathS 0 (canonS lambda j) (canonS lambda i)

Cor12_1 :
forall [alpha : T1], nf alpha ->
      forall (beta : T1) (i n : nat),
      beta t1< alpha ->
     i <= n -> const_pathS i alpha beta -> 
     const_pathS n alpha beta
\end{Coqanswer}

@@

\begin{Coqsrc}
H'_L_ : forall (alpha : E0) (i : nat), H'_ alpha i <= L_ alpha (S i)
\end{Coqsrc}

@@

\begin{Coqsrc}
F_vs_Ack : forall n : nat, 2 <= n -> Ack n n <= F_ omega n
\end{Coqsrc}

@@

\begin{Coqsrc}
Remark m_ge_F_omega : forall k,  F_ omega (S k) <= m (S k).
\end{Coqsrc}

@@

\begin{Coqsrc}
Remark m_dominates_Ack : 
     dominates (fun n =>  S (m n)) (fun n => Ack.Ack n n).
(* ... *)

Remark SmNotPR : isPR 1 (fun n => S (m n)) -> False.
(* ... *)

Theorem LNotPR : False.
  Proof.
    apply SmNotPR,  compose1_1IsPR.
    - apply mIsPR.
    - apply succIsPR.
  Qed.

End battle_lenght_notPR.
\end{Coqsrc}

>>>>

\input{movies/snippets/AckNotPR/majorSucc}

\input{movies/snippets/AckNotPR/majorZero}

\input{movies/snippets/AckNotPR/majorProjection}

The remaining cases are proved within a mutual  induction.

@@

\input{movies/snippets/AckNotPR/majorAnyPR}

@@

\input{movies/snippets/AckNotPR/majorPR2}

@@

\input{movies/snippets/Ack/AckStrictMonoL}

@@

\input{movies/snippets/AckNotPR/majorPR2Strict}

@@

\input{movies/snippets/AckNotPR/AckNotPR}

\begin{remark}
It is easy to prove that any unary function which dominatates \texttt{fun n => Ack n n} fails to be primitive recursive. To this end, we use an instance of \texttt{majorAnyPR} dealing with unary functions.

@@

\input{movies/snippets/AckNotPR/majorPR1}

@@

\input{movies/snippets/AckNotPR/domAckNotPR}

\end{remark}

@@

\input{movies/snippets/Battle_length/BattleLength}

@@

\input{movies/snippets/Hydra_Theorems/battleLengthNotPRa}

@@

\input{movies/snippets/Hydra_Theorems/battleLengthNotPRb}

@@

\input{movies/snippets/Hydra_Theorems/battleLengthNotPRc}

@@

\input{movies/snippets/Hydra_Theorems/battleLengthNotPRd}

@@

In \href{../theories/html/hydras.Epsilon0.F_alpha.html}{Epsilon0.F\_alpha}, we prove a relation between the $F$ and $H'$ functionals. For any $\alpha$ and $k>0$,
$H'_{\omega^\alpha}(k)\geq F_\alpha(k)$.

\input{movies/snippets/F_alpha/HprimeF}

Our proof of this lemma is not trivial at all, it uses some properties of the Ketonen-Solovay's toolkit. We advise the reader to explore this proof, with the help of an ide or softwre like alectryon.
%%
%%% To move the path chapter
%%%%

% \begin{Coqsrc}
%   alpha : E0
%   IHalpha : forall beta : E0, beta o< alpha -> P beta
%   Halpha : Limitb alpha
%   n : nat
%   ============================
%   H'_ (Phi0 (CanonS alpha n)) (S n) <= 
%   H'_ (Phi0 (CanonS alpha (S n))) (S n)
% \end{Coqsrc}

% In mathematical notation: $H'_{\omega^{\canonseq{\alpha}{n}}}(n+1) \leq
% H'_{\omega^{\canonseq{\alpha}{n+1}}}(n+1)$.

% \vspace{4pt}

% But there exists no lemma saying that, if 
% $\beta\leq \alpha$, then $H'_\beta(k)\leq H'_\alpha(k)$, for any $\alpha$ and $\beta$. For instance, 
% $H'_{42}(3)=45> H'_\omega(3)=7$.

% Looking for lemmas of the form $H'_\beta(k)\leq H'_\alpha(k)$, we find this one (from our library
% \href{../theories/html/hydras.Epsilon0.Hprime.html}{Epsilon0.Hprime}):

% \begin{Coqanswer}
% H'_restricted_mono_l : 
%     forall (alpha beta : E0) (n : nat), 
%       Canon_plus n alpha beta -> 
%       H'_ beta n <= H'_ alpha n.
% \end{Coqanswer}

% Thus, it remains to prove that 
% there exists a path from ${\omega^{\canonseq{\alpha}{n+1}}}$
% to ${\omega^{\canonseq{\alpha}{n}}}$ composed of 
% $n+1$-steps.

% Fortunately, the Ketonen-Solovay machinery contains three lemmas which help us to build such a path.

% \begin{Coqanswer}
% KS_thm_2_4_lemma5 :
%   forall [i : nat] [alpha beta : T1],
%   const_pathS i alpha beta ->
%   nf alpha -> alpha <> zero -> 
%   const_pathS i (phi0 alpha) (phi0 beta)

% KS_thm_2_4 :
%   forall [lambda : T1], nf lambda ->limitb lambda ->
%   forall i j : nat, i < j -> 
%    const_pathS 0 (canonS lambda j) (canonS lambda i)

% Cor12_1 :
% forall [alpha : T1], nf alpha ->
%       forall (beta : T1) (i n : nat),
%       beta t1< alpha ->
%      i <= n -> const_pathS i alpha beta -> 
%      const_pathS n alpha beta
% \end{Coqanswer}

@@

\input{movies/snippets/L_alpha/HprimeL}

@@

\input{movies/snippets/F_omega/FVsAck}

@@

\input{movies/snippets/Hydra_Theorems/mGeFOmega}

@@

\input{movies/snippets/Hydra_Theorems/mDominatesAck}

\input{movies/snippets/Hydra_Theorems/SmNotPR}

\input{movies/snippets/Hydra_Theorems/LNotPR}
